---
title: "L10 Joins & Missingness"
subtitle: "Data Science 1 with R (STAT 301-1)"
author: "Allison Kane"

format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
    link-external-newwindow: true

execute:
  warning: false
  
from: markdown+emoji 
---

::: {.callout-tip icon=false}

## Github Repo Link

[Allison Github Link](https://github.com/stat301-1-2023-fall/L10-joins-missingness-akane2460.git)

:::

## Load packages

```{r}
#| label: load-pkgs
#| code-fold: false

# Loading package(s)

library(tidyverse)
library(fueleconomy)

```

## Datasets 

```{r}
#| label: datasets

rankings_2021 <- read_rds("data/rankings_2021.rds")
rankings_2022 <- read_rds("data/rankings_2022.rds")

```


## Exercises

### Exercise 1 

Describe in your own words what the functions `left_join`, `right_join`, `inner_join`, and `full_join` accomplish.

::: {.callout-tip icon="false"}
## Solution

left_join describes the function to merge data, keeping all rows observed in the dataframe to be joined to.

right_join describes the function to merge data that keeps all rows observed in the dataframe to be joined by. 

inner_join describes the function to merge data that keeps only rows that appear in both dataframes. 

full_join describes the function to merge data that keeps all rows from both dataframes. 

:::

### Exercise 2

A relationship between `nycflights13::weather` and `nycflights13::airports` is possible. Describe the relationship using appropriate primary/foreign key vocabulary for relational data.


::: {.callout-tip icon="false"}
## Solution

Airports likely has a primary key `faa` and weather has a foreign key `origin`.

:::

### Exercise 3

What weather conditions make it more likely to see a departure delay?

::: {.callout-tip icon="false"}
## Solution

Weather with poor visibility and wind speeds from 10-20 MPH are more likely to see longer departure delays. Flights with higher windspeeds generally might see cancelations, as the number of flights taking off at 40 MPH is much less. 

:::


### Exercise 4

What does `anti_join(flights, airports, by = c("dest" = "faa"))` tell you? 

What does `anti_join(airports, flights, by = c("faa" = "dest"))` tell you?

::: {.callout-tip icon="false"}
## Solution

This first anti_join is searching the flights dataset for dest observations that do 
not match faa in the airports dataset. 

The second is searching the airports dataset for faa observations that do not 
match dest in the flights dataset.

:::


### Exercise 5

Combine `fueleconomy::vehicles` and `fueleconomy::common` to find the records for only the most common models.

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: ex 05

inner_join(vehicles, common, by = c("model", "make")) |> 
  group_by(model) |> 
  summarize(
    count = n()
  ) |> 
  arrange(desc(count)) |> 
  head(5)

```
The most common model is the Ford F150 Pickup 2WD.
:::


## Case Study 1

### Exploration 1

Use `rankings_2021` and `rankings_2022` to answer the following questions^[This data was obtained from [Times Higher Education](https://www.timeshighereducation.com/rankings/united-states/2022#!/length/-1/sort_by/rank/sort_order/asc/cols/scores) and [College Scorecard](https://collegescorecard.ed.gov/data)]:

a) What types of explicit missingness exists in `rankings_2021` and `rankings_2022`, if any?
b) What types of implicit missingness exists in `rankings_2021` and `rankings_2022`, if any?
c) What is the "key" and is it a primary, compound primary, foreign, or surrogate key?


::: {.callout-tip icon="false"}
## Solution

a) Both datasets are missing values explicitly in the institution's tution_fee, room_board, salary_after_10, resources_score, engagement_score, outcomes_score, and environment_score.

b) No implicit missingness is observed

c) The key is `institution` and it is a primary key.

:::

### Exploration 2

Combine `rankings_2021` and `rankings_2022` to answer the following questions:

a) Which colleges had tuition increase the most from 2021 to 2022?
b) Only considering colleges ranked in the top 400, which college increased the most in ranking from 2021 to 2022?
c) Which colleges were ranked in 2021 but not ranked in 2022? which colleges were ranked in 2022 but not ranked in 2021?


::: {.callout-tip icon="false"}
## Solution

a) The top 5 colleges with the highest tuition increases were Columbia University, Berea College, Emerson College, University of St Thomas and University of Minnesota.

b) The college with biggest increases in ranking from 2021 to 2022 was Hendrix College

c) Westminster College (Missouri) and Marian University (Indiana) were ranked 2021 but not 2022. No colleges were ranked in 2022 but not 2021.

:::


## Case Study 2

Add a new variable called `year` which has the value of `2021` for each observation of `rankings_2021` and the value of `2022` for each observation of `rankings_2022`. Then use `bind_rows()` to combine `rankings_2021` and `rankings_2022`.  

When might you use `bind_rows` or `bind_cols` instead of a `join`?

::: {.callout-tip icon="false"}
## Solution

```{r}
#| label: case study 2
#| echo: false

rankings_2021 <- rankings_2021 |> 
  mutate(
    year = "2021",
    resources_score = as.numeric(resources_score),
    engagement_score = as.numeric(engagement_score),
    outcomes_score = as.numeric(outcomes_score),
    environment_score = as.numeric(environment_score)
  )

rankings_2022 <- rankings_2022 |> 
  mutate(
    year = "2022",
    resources_score = as.numeric(resources_score),
    engagement_score = as.numeric(engagement_score),
    outcomes_score = as.numeric(outcomes_score),
    environment_score = as.numeric(environment_score)
  )

bind_rows(rankings_2022, rankings_2021)
```

It might be better to use bind_rows or bind_cols if one would rather have a side-by-side comparison of the observations across the same columns or rows. A join is more useful to see a merged dataframe based on a common key.

:::



